---
title: "RK_lm_different_timestamps"
author: "Johannes Gruenewald"
date: "2024-11-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r Load Required Libraries}

# Load Required Libraries

# Spatial Data Handling & Processing
library(sf)
library(sp)
library(terra)
library(raster)
library(stars)
library(spdep)

# Spatial Analysis & Interpolation
library(gstat)
library(randomForest)
library(nlme)
library(automap)
library(mgcv)

# Data Manipulation & Wrangling
library(tidyverse)

# Data Visualization
library(ggplot2)
library(ggspatial)
library(scales)
library(gridExtra)
library(corrplot)

# Mapping & Interactive Visualization
library(tmap)

# Additional
library(lattice)
library(car)
```

--------------------------------------------------------------------------------

```{r Load and preprocess data}
# ================================================
# DATA LOADING, CLEANING, AND COVARIATE PREPARATION
# ================================================

# ================================================
# Complete data import and pre-processing

# # Read temperature data for the entire year of 2023
# temp <- sf::read_sf('../data/temp_HD/sensor_data_year_2023.geojson')
# 
# # Removes rows where temp is NA
# temp <- temp %>% filter(!is.na(temperature))
# 
# # Reproject the data for Germany
# temp <- st_transform(temp, crs="EPSG:32632")
# 
# # Clean the data from outliers -> currently working with hard, manual breaks
# # Currently nothing is removed by this step because no measurements are above 40°C for the year 2023
# temp_clean <- temp %>% filter(temperature >= -35, temperature <= 40)
# 
# # Calculate hour count since epoch, rounded down to current hour
# temp_clean$chour <- floor(as.numeric(temp_clean$dateobserved) / 3600)
# 
# # Drop all measurement stations which didn't collect data the entire year around
# temp_clean_station <- temp_clean %>%
#   mutate(dateobserved = as.POSIXct(dateobserved, tz = "UTC")) %>%
#   filter(!entity_id %in% c(
#     "hd:DE_Heidelberg_69120_12:WeatherObserved",
#     "hd:DE_Heidelberg_69120_34:WeatherObserved",
#     "hd:DE_Heidelberg_69123_41:WeatherObserved",
#     "hd:DE_Heidelberg_46:WeatherObserved"
#   ))
# 
# # Check if exactly 4 stations were dropped
# length(unique(na.omit(temp_clean$entity_id)))
# length(unique(na.omit(temp_clean_station$entity_id)))

# Calculate hourly mean temp for cleaned data
# temp_clean_station_hm <- temp_clean_station %>%
#     group_by(entity_id, chour) %>%
#     summarise(hm_temp = mean(temperature, na.rm = TRUE), .groups = "drop") %>%
#     mutate(hour = as.POSIXct(chour * 3600, origin = "1970-01-01", tz = "Europe/Berlin"))

# ================================================
# Import of pre-processed temp data and import and pre-processing of covariates 

# consistent timezone for the whole script
tz_berlin <- "Europe/Berlin"

#Shortcut -> import pre-processed hourly means
temp_clean_station_hm <- sf::read_sf('../data/temp_HD/temp_clean_station_hm.gpkg')

temp_clean_station_hm <- st_transform(temp_clean_station_hm, crs="EPSG:32632")

# Read admin boundaries of HD
hd_bounds <- sf::read_sf('../data/AOI/OSM_boundaries_HD.geojson')

# Reproject
hd_bounds <- st_transform(hd_bounds, crs="EPSG:32632")

# Load the elevation data
dem <- terra::rast("../data/DEM/hd_elevation_4326.tif")

# Reproject the elevation data to match correct CRS
dem <- terra::project(dem, y = crs(temp_clean_station_hm))

# Load the elevation data
tch <- terra::rast("../data/canopy_height/hd_canopy_height_4326.tif")

# Mask out all values which are assigned to water and have value 101
# Not part of this area. But needs to be masked as well: 102 Snow/ice; 103 No data
tch[tch == 101] <- NA

# Reproject the elevation data to match correct CRS
tch <- terra::project(tch, y = crs(temp_clean_station_hm))

# Check if the two covariates have the same crs
crs(tch) == crs(dem)

# Load the elevation data
ghs_bh <- terra::rast("../data/building_height/hd_building_height_4326.tif")

# Reproject the elevation data to match correct CRS
ghs_bh <- terra::project(ghs_bh, y = crs(temp_clean_station_hm))

# Check if CRS is correct
crs(ghs_bh) == crs(dem)

# ================================================
# Create a Raster Stack for all covariate information

# Check resolution of each raster
terra::res(dem)
terra::res(tch)
terra::res(ghs_bh)

# Align TCH and building height to DEM resolution and extent
tch_aligned    <- terra::resample(tch, dem, method = "bilinear")
ghs_bh_aligned <- terra::resample(ghs_bh, dem, method = "near")

# Stack all layers
env_stack <- c(dem, tch_aligned, ghs_bh_aligned)

# Rename layers for clarity
names(env_stack) <- c("elevation", "canopy_height", "building_height")

# ================================================
# Code to include lat and lon information in raster stack

# Also add lat and lon to the raster stack
# Create coordinate rasters
lon <- terra::xFromCell(env_stack, 1:ncell(env_stack))
lat <- terra::yFromCell(env_stack, 1:ncell(env_stack))

# Convert to raster layers
lon_raster <- dem
lat_raster <- dem
values(lon_raster) <- lon
values(lat_raster) <- lat
names(lon_raster) <- "lon"
names(lat_raster) <- "lat"

# Stack the three raster layers
grids <- c(env_stack, lon_raster, lat_raster)

# ================================================
# Conversion when running the analysis without proximity to water
# Preparation of grid as a covariate for running RK

# Convert SpatRaster to RasterLayer (raster package)
grids_ras <- stack(grids)

# Now convert to SpatialGridDataFrame
grids_sp <- as(grids_ras, "SpatialGridDataFrame") 

grids_sp@data

# Extract coordinates of the measurement stations
# maybe use them as covariates

# ================================================
# # Extract coordinates and data into a data frame
# coords <- st_coordinates(temp_clean_station_hm)
# locs <- cbind(temp_clean_station_hm, coords)
# locs <- as.data.frame(locs)
# 
# # Keep only one row per station and select relevant columns
# station_coords <- locs %>%
#   dplyr::distinct(entity_id, .keep_all = TRUE) %>%
#   dplyr::select(entity_id, X, Y)
# 
# coordinates(station_coords) <- ~ X + Y
# proj4string(station_coords) <- CRS(proj4string(grids_sp))
# 
# length(unique(stations$entity_id))

# Shortcut: load the pre-processed layer
station_coords_sf <- sf::read_sf('../data/temp_HD/station_coords.gpkg')

# Reproject
station_coords_sf <- st_transform(station_coords_sf, crs="EPSG:32632")

# Convert to SpatialPointsDataFrame
station_coords <- as(station_coords_sf, "Spatial")
```

--------------------------------------------------------------------------------

```{r create_output_folder}
output_dir <- "../rk_results_07-09-2023"
if (!dir.exists(output_dir)) dir.create(output_dir)

# Define subdirectories for specific output types
variogram_dir <- file.path(output_dir, "variogram")
cv_dir        <- file.path(output_dir, "cross_validation")
rk_dir        <- file.path(output_dir, "rk_predictions")

# Create subdirectories if they don't exist
dir.create(variogram_dir, showWarnings = FALSE)
dir.create(cv_dir, showWarnings = FALSE)
dir.create(rk_dir, showWarnings = FALSE)

# ================================================
# Datetime helpers (labels & filenames)
# ================================================
ts_to_posix <- function(ch) {
  # If already POSIXct, return as is
  if (inherits(ch, "POSIXt")) {
    return(ch)
  }
  # If numeric (hours since 1970)
  if (is.numeric(ch)) {
    return(as.POSIXct(ch * 3600, origin = "1970-01-01", tz = "Europe/Berlin"))
  }
  # If character string, try to parse
  as.POSIXct(ch, format = "%Y-%m-%d %H:%M:%S", tz = "Europe/Berlin")
}

nice_dt     <- function(x) format(x, "%Y-%m-%d %H:00 %Z")                                   
file_dt     <- function(x) format(x, "%Y-%m-%d_%H00")                                       
```


```{r}
temp_clean_station_hm
```

```{r main_loop}
# ================================================
# Loop through 24 hourly time stamps in `temp_clean_station_hm`
# ================================================

# Get unique time stamps
# timestamps <- unique(temp_clean_station_hm$chour)[1:168]  

# ================================================
# Select timestamps for analysis window
# ================================================

timestamps_all <- sort(unique(temp_clean_station_hm$hour))

# Define your analysis period (example: one day in July 2023)
start_date <- as.POSIXct("2023-07-09 00:00:00", tz = "Europe/Berlin")
end_date   <- as.POSIXct("2023-07-09 23:00:00", tz = "Europe/Berlin")

# Keep only timestamps within the desired window
timestamps <- timestamps_all[timestamps_all >= start_date & timestamps_all <= end_date]

# Safety check
cat("Looping from:", min(timestamps), "to", max(timestamps), "\n")

# Counters + progress bar
total     <- length(timestamps)
completed <- 0L
skipped   <- 0L
pb <- txtProgressBar(min = 0, max = total, style = 3)
cat(sprintf("Processing %d timestamps…\n", total))

# Loop through each time stamp
total_metrics <- data.frame()
all_temp_data <- data.frame()

for (i in seq_along(timestamps)) {
  ts <- timestamps[i]                                    
  dt_local <- ts_to_posix(ts)                     
  dt_lab   <- nice_dt(dt_local)                   
  dt_file  <- file_dt(dt_local)                   
  cat(sprintf("\n[%d/%d] Processing: chour=%s  (%s)\n", i, total, ts, dt_lab))  

  # ------------------------------------------------
  # STEP 1: Filter hourly data and plot temperature
  # ------------------------------------------------
  
  # Select temperature data for the current hour
  sel_temp_hm <- temp_clean_station_hm %>%
    filter(hour == ts)
  
  # Skip if fewer than 20 data points for current hour
  if (nrow(sel_temp_hm) < 20) {                                 
    skipped <- skipped + 1L                                       
    cat("  → skipped (n < 20)\n")                          
    next                                                         
  }
  
  # Append the selected temperature data to full dataset
  all_temp_data <- rbind(all_temp_data, sel_temp_hm)
  
  # Generate line plot showing hourly mean temperature for each station
  # across the time range covered so far in the loop
  p_all <- ggplot(all_temp_data, aes(x = hour, y = hm_temp, color = entity_id, group = entity_id)) +  
    geom_line() +
    geom_point(size = 1) +
    labs(title = "Hourly Mean Temperature at Each Station",        
         x = "Date & Hour (Europe/Berlin)", y = "Mean Temperature (°C)", color = "Station") + 
    scale_x_datetime(date_labels = "%d.%m\n%H:%M", date_breaks = "12 hours") +              
    theme_minimal(base_size = 12) +
    theme(axis.text.x = element_text(angle = 0, hjust = 0.5))                                  
  
  # Save the temperature plot
  ggsave(file.path(output_dir, "all_stations_hourly_mean_temperature.png"),
         plot = p_all, width = 10, height = 6, dpi = 300)

  # ------------------------------------------------
  # STEP 2: Add covariates (elev, canopy, bldg) from grid
  # ------------------------------------------------
  
  # Convert Spatial object to RasterBrick
  r_brick <- brick(grids_sp)
  
  # Convert RasterBrick to terra SpatRaster object
  grid_terra <- rast(r_brick)
  
  # Extract raster values at the point locations
  extracted_vals <- terra::extract(grid_terra, sel_temp_hm)
  
  # Remove the first column (ID)
  extracted_vals <- extracted_vals[,-1]
  
  # Combine the point data with the extracted raster values
  temp_shm.ex <- cbind(as.data.frame(sel_temp_hm), extracted_vals)
  
  # ------------------------------------------------
  # STEP 3: Fit linear model and prepare spatial data
  # ------------------------------------------------
  
  # Fit multiple linear regression model
  lm.temp_shm <- lm(hm_temp ~ elevation + canopy_height + building_height, data = temp_shm.ex)
  summary_lm <- summary(lm.temp_shm)
  
  # Compute residuals and predictions
  pred <- predict(lm.temp_shm, temp_shm.ex)
  resid <- temp_shm.ex$hm_temp - pred
  
  # Extract metrics
  model_metrics <- data.frame(
    timestamp      = ts,
    r_squared      = summary_lm$r.squared,
    adj_r_squared  = summary_lm$adj.r.squared,
    f_statistic    = summary_lm$fstatistic[1],
    f_pvalue       = pf(summary_lm$fstatistic[1],
                        summary_lm$fstatistic[2],
                        summary_lm$fstatistic[3],
                        lower.tail = FALSE),
    residual_se    = summary_lm$sigma,
    rmse           = sqrt(mean(resid^2)),
    mae            = mean(abs(resid)),
    AIC            = AIC(lm.temp_shm),
    BIC            = BIC(lm.temp_shm)
  )
  
  # Append to tracking data frame
  if (!exists("lm_metrics_all")) {
    lm_metrics_all <- model_metrics
  } else {
    lm_metrics_all <- rbind(lm_metrics_all, model_metrics)
  }

  # Save model's residuals back into original df as new column
  temp_shm.ex$residuals <- resid
  
  # Convert df into spatial object using existing geometry and
  # Convert MULTIPOINT geometries to POINT
  temp_shm.sf <- st_as_sf(temp_shm.ex) |> st_cast("POINT")
  
  st_crs(temp_shm.sf) <- st_crs(grids_sp)
  
  # ------------------------------------------------
  # STEP 4: Compute anisotropy & fit directional variogram
  # ------------------------------------------------
  
  # Compute empirical variogram map of model residuals
  # This helps visualize spatial autocorrelation in all directions
  varmap <- variogram(residuals ~ 1, data = temp_shm.sf, map = TRUE,
                    # cutoff is half the spatial extent (or diameter) of the study area
                    cutoff = sqrt(areaSpatialGrid(grids_sp)) / 2, 
                    # scaled relative to the resolution of the grid
                    width = 30 * grids_sp@grid@cellsize[1])

  # Store cutoff and bin width values for reuse and labeling
  cutoff_val <- sqrt(areaSpatialGrid(grids_sp)) / 2
  width_val  <- 30 * grids_sp@grid@cellsize[1]
  
  # Report how many rows are in variogram map
  cat("Variogram map has", nrow(as.data.frame(varmap)), "rows\n")
  
  # Define output filename for the variogram map plot
  file_varmap <- file.path(variogram_dir, paste0(dt_file, "_variogram_map.png"))  
  
  # Plot and save the directional variogram map
  png(file_varmap, width = 8, height = 6, units = "in", res = 300)
  print(
    plot(varmap,
         col.regions = grey(rev(seq(0, 1, 0.025))),
         main = sprintf("Directional Variogram Map (cutoff: %.1f km, width: %.1f km)\n%s",  
                        cutoff_val / 1000, width_val / 1000, dt_lab))
  )
  dev.off()
  
  # Compute directional variograms for North–South (0°) and East–West (90°)
  rv.temp <- variogram(residuals ~ 1, temp_shm.sf, alpha = c(0, 45, 90, 135))

  # Fit theoretical exponential variogram model to the directional variogram
  # Compute experimental variogram
  rv.temp <- variogram(residuals ~ 1, temp_shm.sf)
  
  # Try several initial models and pick the best fit
  init_models <- list(
    vgm("Sph"),
    vgm("Exp"),
    vgm("Gau")
  )
  
  fit_list <- lapply(init_models, function(m) fit.variogram(rv.temp, m))
  fit_errors <- sapply(fit_list, attr, "SSErr")
  rvgm.temp <- fit_list[[which.min(fit_errors)]]

  
  # Define output filename for directional variogram fit plot
  file_varfit <- file.path(variogram_dir, paste0(dt_file, "_directional_variogram_fit.png"))  
 
  # Plot the empirical and fitted directional variograms and save
  png(file_varfit, width = 8, height = 6, units = "in", res = 300)
  print(
    plot(rv.temp, rvgm.temp,
         main = paste0("Directional Variogram Fit (0°, 45°, 90°, 135°)\n", dt_lab),   
         plot.nu = FALSE, cex = 1.5, pch = "+", col = "black")
  )
  dev.off()
  
  # ------------------------------------------------
  # STEP 5: Cross-validation of residual kriging model
  # ------------------------------------------------

  # Perform leave-one-out cross-validation (LOOCV) of the kriging model
  cv_resid <- krige.cv(residuals ~ 1,
                       locations = temp_shm.sf,
                       model = rvgm.temp,
                       nfold = nrow(temp_shm.sf),   # one fold per observation = LOOCV
                       nmax = 10)                   # max number of neighbors used for prediction
  
  # Compute CV metrics
  cv_rmse <- sqrt(mean(cv_resid$residual^2))
  cv_mae  <- mean(abs(cv_resid$residual))
  cv_me   <- mean(cv_resid$residual)
  
  cat("CV RMSE:", round(cv_rmse, 2), "\n")
  cat("CV MAE :", round(cv_mae, 2), "\n")
  cat("CV Bias:", round(cv_me, 2), "\n")
  
  # Create df with CV metrics for current timestamp
  cv_metrics <- data.frame(
    timestamp = ts,
    cv_rmse   = cv_rmse,
    cv_mae    = cv_mae,
    cv_bias   = cv_me
  )
  
  # Append metrics to tracking df across all time steps
  if (!exists("cv_metrics_all")) {
    cv_metrics_all <- cv_metrics
  } else {
    cv_metrics_all <- rbind(cv_metrics_all, cv_metrics)
  }
  
  # Create and save scatterplot of predicted vs residual values from CV
  p_cv <- ggplot(cv_resid, aes(x = var1.pred, y = residual)) +
    geom_point() +
    geom_hline(yintercept = 0, linetype = "dashed") +
    labs(title = paste("Cross-validation Residuals\n", dt_lab), x = "Predicted", y = "Residual") +  
    theme_minimal(base_size = 12)
  
  ggsave(filename = file.path(cv_dir, paste0(dt_file, "_cv_residuals_plot.png")),  
         plot = p_cv, width = 8, height = 6, dpi = 300)

  # ------------------------------------------------
  # STEP 6: Predict regression trend and kriging residuals
  # ------------------------------------------------
    
  # Define a gstat object for kriging using the fitted residual variogram
  g.resid <- gstat(
    id = c("residuals"),
    formula = residuals ~ 1,
    data = temp_shm.sf,
    nmax = 31,
    model = rvgm.temp
  )
  
  # Predict the temperature trend component using the regression model
  locTEMP.reg <- predict(lm.temp_shm, grids_sp)
  
  # OK of the regression residuals to model spatial autocorrelation
  locTEMP <- predict(g.resid, grids_sp, beta=1, BLUE=FALSE)
  
  # Combine the regression prediction with kriged residuals to get final RK predictions
  rk.pred <- locTEMP.reg + locTEMP$residuals.pred
  
  # Add the RK predictions as a new variable to the spatial grid object
  grid_out <- grids_sp
  grids_sp$rk_pred <- rk.pred
  
  # Convert SpatialPixelsDataFrame → SpatRaster for export
  rk_raster <- terra::rast(grids_sp["rk_pred"])
  
  # Define output .tif path
  rk_file_tif <- file.path(rk_dir, paste0(dt_file, "_rk_prediction_map.tif"))
  
  # Write GeoTIFF
  terra::writeRaster(
    rk_raster,
    filename = rk_file_tif,
    overwrite = TRUE,
    gdal = c("COMPRESS=LZW")  # optional compression
  )
  
  cat("  → RK prediction saved as GeoTIFF:", rk_file_tif, "\n")
  
  # Define output filename for the RK prediction map
  rk_file_img <- file.path(rk_dir, paste0(dt_file, "_rk_prediction_map", ".png"))   
  
  # Generate and save the RK prediction map
  png(rk_file_img, width = 8, height = 6, units = "in", res = 300)
  print(
    spplot(grids_sp, "rk_pred",
           main = paste("Regression Kriging Prediction (°C)\n", dt_lab),   
           sp.layout = list("sp.points", station_coords, pch = 20, col = "white"))
  )
  dev.off()
  
  # ------------------------------------------------
  # STEP 7: Evaluate RK predictions at observation stations
  # ------------------------------------------------
  
  # Match station coordinates with observed stations only
  obs_coords <- st_coordinates(temp_shm.sf)
  
  # Convert to SpatialPoints object with same CRS as prediction grid
  obs_stations <- SpatialPoints(coords = obs_coords, proj4string = CRS(proj4string(grids_sp)))
  
  # Extract only at observed station locations
  rk_at_obs <- sp::over(obs_stations, grids_sp)
  
  # Combine predictions with observations
  rk_eval <- cbind(rk_at_obs, obs = temp_shm.sf$hm_temp)

  # Check if predicted and observed data match in length
  cat("Predicted rows:", nrow(rk_at_obs), "\n")
  cat("Observed rows :", length(temp_shm.sf$hm_temp), "\n")
  
  # Calculate prediction error
  rk_eval$residual <- rk_eval$obs - rk_eval$rk_pred
  
  # Calculate evaluation metrics for RK performance
  rmse <- sqrt(mean(rk_eval$residual^2, na.rm = TRUE))
  mae  <- mean(abs(rk_eval$residual), na.rm = TRUE)
  bias <- mean(rk_eval$residual, na.rm = TRUE)
  
  cat("RK RMSE :", round(rmse, 2), "\n")
  cat("RK MAE  :", round(mae, 2), "\n")
  cat("RK Bias :", round(bias, 2), "\n")
  
  # Store metrics in df for current timestamp
  rk_metrics <- data.frame(
    timestamp = ts,
    rk_rmse   = rmse,
    rk_mae    = mae,
    rk_bias   = bias
  )
  
  # Append metrics to cumulative RK metrics tracking object
  if (!exists("rk_metrics_all")) {
    rk_metrics_all <- rk_metrics
  } else {
    rk_metrics_all <- rbind(rk_metrics_all, rk_metrics)
  }
  
  # Save RK residuals plot
  rk_resid_plot <- ggplot(rk_eval, aes(x = rk_pred, y = residual)) +
    geom_point() +
    geom_hline(yintercept = 0, linetype = "dashed") +
    labs(title = paste("RK Prediction Residuals at Stations\n", dt_lab),   
         x = "RK Predicted (°C)", y = "Residual (Observed - Predicted)") +
    theme_minimal(base_size = 12)
  
  ggsave(filename = file.path(rk_dir, paste0(dt_file, "_rk_residuals_plot.png")),  
         plot = rk_resid_plot, width = 8, height = 6, dpi = 300)
  
  completed <- completed + 1L 
  setTxtProgressBar(pb, completed)
  cat("\n")
  cat(sprintf("  → done. Completed: %d/%d (skipped: %d)\n",      
              completed, total, skipped))                        
}

# Close progress bar + final summary
close(pb)                                                        
cat(sprintf("\nFinished. Completed: %d/%d | Skipped: %d\n",       
            completed, total, skipped))                         

# Save all LM metrics to a single CSV file
metrics_file <- file.path(output_dir, "lm_model_performance_summary.csv")
if (exists("lm_metrics_all")) {                                     
  lm_metrics_all$datetime_local <- ts_to_posix(lm_metrics_all$timestamp)   
  lm_metrics_all <- lm_metrics_all %>% relocate(datetime_local, .after = timestamp)  
  write.csv(lm_metrics_all, metrics_file, row.names = FALSE)
}

# Save all CV metrics to a single CSV file
cv_file <- file.path(output_dir, "cross_validation_metrics.csv")
if (exists("cv_metrics_all")) {                                     
  cv_metrics_all$datetime_local <- ts_to_posix(cv_metrics_all$timestamp) 
  cv_metrics_all <- cv_metrics_all %>% relocate(datetime_local, .after = timestamp) 
  write.csv(cv_metrics_all, cv_file, row.names = FALSE)
}

# Save all RK metrics to a single CSV file
rk_file <- file.path(output_dir, "rk_metrics.csv")
if (exists("rk_metrics_all")) {                                   
  rk_metrics_all$datetime_local <- ts_to_posix(rk_metrics_all$timestamp)
  rk_metrics_all <- rk_metrics_all %>% relocate(datetime_local, .after = timestamp) 
  write.csv(rk_metrics_all, rk_file, row.names = FALSE)
}
```

------------------------------------------------------------------------

Plots to analyse model performance metrices

```{r Analyse CV metrics}
cv_metrics <- read.csv("../rk_results_07-09-2023/cross_validation_metrics.csv",
                       stringsAsFactors = FALSE)

cv_metrics <- cv_metrics %>%
  mutate(
    # ✅ Explicitly parse with full datetime format
    datetime_calc = as.POSIXct(datetime_local, format = "%Y-%m-%d %H:%M:%S", tz = "Europe/Berlin"),
    date = as.Date(datetime_calc),
    hour = lubridate::hour(datetime_calc)
  )

# Reshape to long format
cv_metrics_long <- cv_metrics %>%
  pivot_longer(cols = c(cv_rmse, cv_mae),
               names_to = "metric",
               values_to = "value")

# Filter nur 2023
df_plot <- cv_metrics_long %>%
  filter(year(date) == 2023)

df_plot$hour <- as.numeric(df_plot$hour)

# Plot: Daily Evaluation of the Regression-Kriging Cross-Validation Metrics
p_cv <- ggplot(df_plot, aes(x = hour, y = value, color = metric, group = metric)) +
  # Lines, reference line, and points
  geom_line(linewidth = 1.2) +
  geom_hline(
    aes(yintercept = 0, linetype = "Perfect Prediction (Error = 0)"),
    color = "grey60"
  ) +
  geom_point(size = 1.8) +
  # Facet by date
  facet_wrap(~ date, ncol = 3, scales = "fixed") +   # gleiche y-Achse
  # Color and linetype scales
  scale_color_manual(
    values = c(
      "cv_rmse" = "#6699CC",
      "cv_mae"  = "#CC6677"
    ),
    labels = c(
      "cv_rmse" = "Root Mean Square Error (RMSE)",
      "cv_mae"  = "Mean Absolute Error (MAE)"
    )
  ) +
  scale_linetype_manual(
    name   = "",
    values = c("Perfect Prediction (Error = 0)" = "dotted")
  ) +
  # Labels
  labs(
    title    = "Daily Evaluation of the Regression-Kriging Leave-One-Out Cross-Validation Metrics",
    subtitle = "Hourly RMSE and MAE values from LOOCV of models predicting hourly mean temperature",
    x        = "Hour of the Day",
    y        = "Error Metric [°C]",
    color    = "Metrics"
  ) +
  # X-axis breaks
  scale_x_continuous(breaks = c(0, 6, 12, 18, 24)) +
  # Theme
  theme_minimal(base_size = 14) +
  theme(
    legend.position   = "top",
    panel.grid.minor  = element_blank(),
    strip.background  = element_rect(fill = "grey90", color = NA),
    strip.text        = element_text(face = "bold")
  )

p_cv

# Save the plot
#ggsave("../plots/cv_metrics_by_day_01-01-2023.png", plot = p_cv, width = 12, height = 8, dpi = 300)
```

```{r Analyse RK metrics}

# Read CSV
rk_metrics <- read.csv("../rk_results_07-09-2023/rk_metrics.csv",
                       stringsAsFactors = FALSE)

# Parse datetime correctly (same as CV metrics)
rk_metrics <- rk_metrics %>%
  mutate(
    datetime_calc = as.POSIXct(datetime_local, format = "%Y-%m-%d %H:%M:%S", tz = "Europe/Berlin"),
    date = as.Date(datetime_calc),
    hour = lubridate::hour(datetime_calc)
  )

# Reshape to long format
rk_metrics_long <- rk_metrics %>%
  pivot_longer(cols = c(rk_rmse, rk_mae),
               names_to = "metric",
               values_to = "value")

# Filter year 2023
df_plot <- rk_metrics_long %>%
  filter(year(date) == 2023)

# Make sure hour is numeric
df_plot$hour <- as.numeric(df_plot$hour)

# Plot: Daily Evaluation of the Regression-Kriging Model Performance Metrics
p_rk <- ggplot(df_plot, aes(x = hour, y = value, color = metric, group = metric)) +
  # Lines, reference line, and points
  geom_line(linewidth = 1.2) +
  geom_hline(
    aes(yintercept = 0, linetype = "Perfect Prediction (Error = 0)"),
    color = "grey60"
  ) +
  geom_point(size = 1.8) +
  # Facet by date
  facet_wrap(~ date, ncol = 3, scales = "fixed") +   # gleiche y-Achse
  # Color and linetype scales
  scale_color_manual(
    values = c(
      "rk_rmse" = "#6699CC",
      "rk_mae"  = "#CC6677"
    ),
    labels = c(
      "rk_rmse" = "Root Mean Square Error (RMSE)",
      "rk_mae"  = "Mean Absolute Error (MAE)"
    )
  ) +
  scale_linetype_manual(
    name   = "",
    values = c("Perfect Prediction (Error = 0)" = "dotted")
  ) +
  # Labels
  labs(
    title    = "Daily Evaluation of the Regression-Kriging Model Performance Metrics",
    subtitle = "Hourly RMSE and MAE values from models predicting hourly mean temperature",
    x        = "Hour of the Day",
    y        = "Error Metric [°C]",
    color    = "Metrics"
  ) +
  # X-axis breaks
  scale_x_continuous(breaks = c(0, 6, 12, 18, 24)) +
  # Theme
  theme_minimal(base_size = 14) +
  theme(
    legend.position   = "top",
    panel.grid.minor  = element_blank(),
    strip.background  = element_rect(fill = "grey90", color = NA),
    strip.text        = element_text(face = "bold")
  )

p_rk

# Save plot
# ggsave("../plots/rk_metrics_week-01-2023.png", plot = p_rk, width = 12, height = 8, dpi = 300)
```

```{r Analyse LM metrics}

# Read CSV
lm_metrics <- read.csv("../rk_results_07-09-2023/lm_model_performance_summary.csv",
                       stringsAsFactors = FALSE)

# Parse datetime column correctly (same approach as CV & RK)
lm_metrics <- lm_metrics %>%
  mutate(
    datetime_calc = as.POSIXct(datetime_local, format = "%Y-%m-%d %H:%M:%S", tz = "Europe/Berlin"),
    date = as.Date(datetime_calc),
    hour = lubridate::hour(datetime_calc)
  )

# Reshape to long format
lm_metrics_long <- lm_metrics %>%
  pivot_longer(cols = c(r_squared, adj_r_squared),
               names_to = "metric",
               values_to = "value")

# Filter year 2023
lm_df_plot <- lm_metrics_long %>%
  filter(year(date) == 2023)

# Make sure hour is numeric
lm_df_plot$hour <- as.numeric(lm_df_plot$hour)

# Plot: Daily Evaluation of Multiple Linear Regression Model Performance
p_lm <- ggplot(lm_df_plot, aes(x = hour, y = value, color = metric, group = metric)) +
  # Lines and points for model metrics
  geom_line(linewidth = 1.2) +
  geom_hline(
    aes(yintercept = 1, linetype = "Perfect Fit (R² = 1)"),
    color = "grey60"
  ) +
  geom_point(size = 1.8) +
  # Facet by date
  facet_wrap(~ date, ncol = 3, scales = "fixed") +
  # Color and linetype scales
  scale_color_manual(
    values = c(
      "r_squared"     = "#6699CC",
      "adj_r_squared" = "#CC6677"
    ),
    labels = c(
      "r_squared"     = expression(R^2),
      "adj_r_squared" = expression(Adjusted~R^2)
    )
  ) +
  scale_linetype_manual(
    name   = "",
    values = c("Perfect Fit (R² = 1)" = "dotted")
  ) +
  # Labels
  labs(
    title    = "Daily Evaluation of Multiple Linear Regression Model Performance",
    subtitle = "Hourly R² and Adjusted R² values from models predicting hourly mean temperature",
    x        = "Hour of the Day",
    y        = "Metric Value",
    color    = "Metrics"
  ) +
  # X-axis ticks
  scale_x_continuous(breaks = c(0, 6, 12, 18, 24)) +
  # Theme
  theme_minimal(base_size = 14) +
  theme(
    legend.position    = "top",
    legend.box         = "horizontal",
    panel.grid.minor   = element_blank(),
    panel.spacing      = unit(0.8, "lines"),
    strip.background   = element_rect(fill = "grey90", color = NA),
    strip.text         = element_text(face = "bold")
  )

# Show plot
p_lm

# Save plot
# ggsave("../plots/lm_metrics_by_day_01-01-2023.png", plot = p_lm, width = 12, height = 8, dpi = 300)
```


------------------------------------------------------------------------

CURRENTLY EXCLUDED. IS PROXIMITY TO WATER COVARIATE

Can be included, but destroys the results -> especially the visual appearance

```{r}
# Add proximity to water as another covariate

# Convert grid data to point features
points_vect <- as.points(grids)

# Convert point vector to sf object for spatial analysis
grid_sf <- st_as_sf(points_vect)

# Load river network
neckar <- st_read("../data/waterways/waterways_HD.gpkg")

# Reproject to 32632
neckar <- st_transform(neckar, crs = st_crs(temp))

# Clip Neckar river to extent of the grid (AOI)
neckar_clipped <- st_intersection(neckar, st_as_sfc(st_bbox(grid_sf)))

# Calculate minimum distance from each grid point to the Neckar river
grid_sf$dwater <- st_distance(grid_sf, neckar_clipped) %>%
  apply(1, min) %>%
  as.numeric()

# Cap at 500 meters to limit influence of distant points
#grid_sf$dwater[grid_sf$dwater > 500] <- NA

# Move 'dist_to_neckar' to first column
grid_sf <- grid_sf[, c("dwater", setdiff(names(grid_sf), "dwater"))]
```

--------------------------------------------------------------------------------

Comparison of the Urban Climate Analysis of HD (2023) with the 
RK temperature prediction surface

```{r Compare the Stadtklima-Analyse with my RK prediction surface}

# Load rasters and vector data
mod  <- rast("../rk_results_07-09-2023/rk_predictions/2023-07-09_1400_rk_prediction_map.tif")
ref5 <- rast("../data/reference/2_22_074_ist_t14_rev01.tif")
river <- st_read("../data/waterways/waterways_HD.gpkg")                 
landuse <- st_read("../data/DE522L1_HEIDELBERG_UA2018_v013/Data/DE522L1_HEIDELBERG_UA2018_v013.gpkg")

# Check if model has CRS
if (is.na(crs(mod))) stop("`mod` has no CRS. Set it first.")

# Define CRS and project reference raster to model grid
crs_mod <- crs(mod, proj = TRUE)
crs(ref5) <- "EPSG:32632"
ref30 <- project(ref5, mod, method = "bilinear")

# Reproject vector layers to project CRS
crs_mod <- crs(mod, proj = TRUE)
hd_bounds <- st_transform(hd_bounds, crs = crs_mod)
river     <- st_transform(river,     crs = crs_mod)
landuse   <- st_transform(landuse,   crs = crs_mod)

# Crop and mask rasters by AOI
mod    <- mask(crop(mod,    vect(hd_bounds)), vect(hd_bounds))
ref30  <- mask(crop(ref30,  vect(hd_bounds)), vect(hd_bounds))

# Clip vector layers to AOI
river <- st_intersection(river, hd_bounds)
landuse <- st_intersection(landuse, hd_bounds)

# Verify alignment between datasets
stopifnot(terra::same.crs(ref30, mod))
print(st_crs(hd_bounds)$input)
stopifnot(isTRUE(all.equal(res(mod),  res(ref30))))
stopifnot(isTRUE(all.equal(ext(mod),  ext(ref30))))
```

```{r Model–Reference Temperature Comparison and Metrics}

# Select single-hour rasters
mod_h   <- mod    
ref30_h <- ref30       

# Apply common NA mask
both_mask <- !is.na(values(mod_h)) & !is.na(values(ref30_h))
mod_h[!both_mask]   <- NA
ref30_h[!both_mask] <- NA

# Extract cell values from both rasters
vr <- values(c(mod_h, ref30_h))           
colnames(vr) <- c("rk_temp", "ref_temp")  # rename columns
comparison_df <- as.data.frame(vr) |> drop_na() 

# Calculate comparison metrics
pearson_r <- cor(comparison_df$rk_temp, comparison_df$ref_temp, method = "pearson")
spearman_r<- cor(comparison_df$rk_temp, comparison_df$ref_temp, method = "spearman")
bias_C    <- mean(comparison_df$rk_temp - comparison_df$ref_temp)
mae_C     <- mean(abs(comparison_df$rk_temp - comparison_df$ref_temp))
rmse_C    <- sqrt(mean((comparison_df$rk_temp - comparison_df$ref_temp)^2))

# Combine metrics for output
metrics <- c(
  pearson_r = pearson_r,
  spearman_r = spearman_r,
  bias_C = bias_C,
  MAE_C = mae_C,
  RMSE_C = rmse_C
)
print(metrics)

# Sample data if very large (for faster plotting)
set.seed(1)
comparison_plot_df <- if (nrow(comparison_df) > 1e6) {
  comparison_df[sample.int(nrow(comparison_df), 1e6), ]
} else comparison_df

# Plot model vs. reference scatter with 1:1 line
ggplot(comparison_plot_df, aes(x = ref_temp, y = rk_temp)) +
  geom_hex(bins = 40) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  coord_equal() +
  labs(
    x = "Reference (5 m downsampled to 30 m mean) [°C]",
    y = "RK model (30 m) [°C]",
    title = sprintf("Agreement at 14:00 — r = %.2f, RMSE = %.2f °C", pearson_r, rmse_C)
  )
```

Short interpretation of the plot:
The RK model reproduces the relative spatial pattern of heat well (r ≈ 0.88) but is systematically too warm by about 8–9 °C across the city at 14:00.

```{r RK and Reference Temperature Map Visualization}

# Convert rasters to data frames for plotting  
mod_df  <- as.data.frame(mod_h,  xy = TRUE, na.rm = TRUE)
ref_df  <- as.data.frame(ref30_h, xy = TRUE, na.rm = TRUE)
names(mod_df)[3] <- "value"
names(ref_df)[3] <- "value"

# Compute shared color scale range  
mm_mod <- terra::minmax(mod_h)
mm_ref <- terra::minmax(ref30_h)
rng <- range(c(mm_mod, mm_ref), na.rm = TRUE)

# Convert vector layers to sf  
river_sf     <- st_as_sf(river)
hd_bounds_sf <- st_as_sf(hd_bounds)

# Plot RK model map  
p_rk_map <- ggplot() +
  geom_raster(data = mod_df, aes(x = x, y = y, fill = value)) +
  geom_sf(data = river_sf, color = "lightblue", linewidth = 0.4) +
  geom_sf(data = hd_bounds_sf, fill = NA, color = "black", linewidth = 0.4) +
  scale_fill_viridis_c(name = "Temperature [°C]", limits = rng) +
  coord_sf() + 
  annotation_scale(location = "br", width_hint = 0.3) +    
  annotation_north_arrow(location = "tl", which_north = "true", width = unit(0.9, "cm"),
                         style = north_arrow_orienteering) +
  labs(title = "RK model",
      subtitle = sprintf("Temperature value range: %.1f – %.1f °C", 
                         min(mod_df$value, na.rm = TRUE), max(mod_df$value, na.rm = TRUE))) +
  theme_minimal(base_size = 13) +
  theme(
    axis.title = element_blank(),
    axis.text  = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 9, face = "plain", 
                                 margin = margin(t = 1, b = 5))
  )

# Plot reference map  
p_ref_map <- ggplot() +
  geom_raster(data = ref_df, aes(x = x, y = y, fill = value)) +
  geom_sf(data = river_sf, color = "lightblue", linewidth = 0.4) +
  geom_sf(data = hd_bounds_sf, fill = NA, color = "black", linewidth = 0.4) +
  scale_fill_viridis_c(name = "Temperature [°C]", limits = rng) +
  coord_sf() +
  annotation_scale(location = "br", width_hint = 0.3) +
  annotation_north_arrow(location = "tl", which_north = "true", width = unit(0.9, "cm"),
                         style = north_arrow_orienteering) +
  labs(title = "Reference model",
       subtitle = sprintf("Temperature value range: %.1f – %.1f °C", 
                          min(ref_df$value, na.rm = TRUE), max(ref_df$value, na.rm = TRUE))) +
  theme_minimal(base_size = 13) +
  theme(
    axis.title = element_blank(),
    axis.text  = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 9, face = "plain", 
                                 margin = margin(t = 1, b = 5))
  )

# Combine both plots with shared legend  
p_rk_ref <- (p_rk_map + p_ref_map + 
                    plot_layout(guides = "collect") & 
                    theme(legend.position = "right")) +
  plot_annotation(
    title = "Comparison of RK and Reference Temperature Maps — Heidelberg, 09-07-2023 14:00",
    theme = theme(plot.title = element_text(size = 16, face = "bold", hjust = 0.5,
        margin = margin(b = 10)))
  )

# Display combined plot
p_rk_ref

# Save as PNG
# ggsave("../plots/rk_vs_ref_14-00.png", p_rk_ref, width = 10, height = 5, dpi = 300)
```

Short explanation of the following chunk:
The following chunk is a quality-control checkpoint. It doesn’t change the analysis output, but it ensures that what I compare is valid (same units, same grid, no hidden bias).
Keep it as a safety check but might omit it from the final report.

```{r Residual diagnostics and consistency checks}

# Residual summary (model-reference)
resid <- comparison_df$rk_temp - comparison_df$ref_temp
summary(resid)
mean(resid)        # mean bias
median(resid)      # median bias
mean(resid < 0)    # share of negative residuals
mean(resid > 0)    # share of positive residuals

# Unit consistency check
summary(comparison_df$rk_temp)
summary(comparison_df$ref_temp)
# Values ~280–320 → Kelvin; ~10–45 → Celsius
# A ~273.15 offset would indicate mixed units

# Spatial alignment check (CRS, resolution, extent)
stopifnot(terra::same.crs(ref30, mod))
stopifnot(isTRUE(all.equal(terra::res(ref30),  terra::res(mod))))
stopifnot(isTRUE(all.equal(terra::ext(ref30),  terra::ext(mod))))
```

Short interpretation of the residual summary metrics:
The RK model systematically overestimates temperatures by about +8.7 °C across all grid cells, with no negative residuals. Both datasets are in °C and perfectly aligned spatially.

```{r Spatial Residual Map and Histogram}

# Residual histogram
p_resid_hist <- ggplot(data.frame(resid), aes(x = resid)) +
  geom_histogram(bins = 60, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 0, color = "black", linetype = "dashed", linewidth = 0.6) +
  labs(
    title = "Residual Distribution (RK Model – Reference, 09-07-2023 14:00)",
    subtitle = sprintf("Mean bias = %.2f °C", mean(resid)),
    x = "Residual (Model − Reference) [°C]",
    y = "Number of cells"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10, margin = margin(b = 5)),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "grey85"),
    axis.title = element_text(face = "bold")
  )

# Recalculate spatial residuals (per grid cell) to retain geographic structure 
resid_raster <- mod_h - ref30_h

# Convert residual raster to data frame for ggplot visualization
resid_map_df <- as.data.frame(resid_raster, xy = TRUE, na.rm = TRUE)
names(resid_map_df)[3] <- "residual"

# Define symmetric color scale limits around zero
rmax <- ceiling(max(abs(terra::global(resid_raster, "min", na.rm = TRUE)$min),
                    abs(terra::global(resid_raster, "max", na.rm = TRUE)$max)))
# round to nearest multiple of 5 for clean legend breaks
rmax <- ceiling(rmax / 5) * 5  

# Plot spatial residual distribution
p_resid_map <- ggplot() +
  geom_raster(data = resid_map_df, aes(x = x, y = y, fill = residual)) +
  geom_sf(data = river_sf, color = "lightblue", linewidth = 0.4) +
  geom_sf(data = hd_bounds_sf, fill = NA, color = "black", linewidth = 0.5) +
  scale_fill_gradient2(
    name = "Residual (°C)",
    low = "#2166ac", mid = "white", high = "#b2182b",
    midpoint = 0,
    limits = c(-rmax, rmax),
    breaks = seq(-15, 15, 5),
    labels = seq(-15, 15, 5)
  ) +
  coord_sf() +
  annotation_scale(location = "br", width_hint = 0.3) +    
  annotation_north_arrow(location = "tl", which_north = "true", width = unit(0.9, "cm"),
                         style = north_arrow_orienteering) +
  labs(
    title = "Spatial Distribution of Residuals (Model − Reference) on 09-07-2023 at 14:00 [°C]",
    subtitle = "Blue: Model colder than reference | Red: Model warmer than reference"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 9, margin = margin(b = 8)),
    legend.position = "right",
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 9),
    axis.text = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank()
  )

# Display the plots
p_resid_hist
p_resid_map

# Save as PNG
ggsave("../plots/residual_histogram_rk_vs_reference_2023-07-09_14-00.png", p_resid_hist, width = 10, height = 5, dpi = 300)

ggsave("../plots/residual_spatial_map_rk_vs_reference_2023-07-09_14-00.png", p_resid_map, width = 10, height = 5, dpi = 300)
```

```{r Robust Performance Metrics and Regression Fit}

# Robust central tendency
median_bias <- median(resid, na.rm = TRUE)

# Linear regression between reference and model temperatures
lm_rk_ref_temp <- lm(ref_temp ~ rk_temp, data = comparison_df)
summary(lm_rk_ref_temp)$coefficients 
summary(lm_rk_ref_temp)$r.squared    

# Print summary results
c(median_bias = median_bias, R2 = summary(lm_rk_ref_temp)$r.squared)
```

Short interpretation of the linear regression model results:
The analysis shows that the RK model reproduces the general spatial temperature pattern well (R² = 0.78) but consistently predicts higher absolute temperatures than the reference. With a median bias of about +8.7 °C (in this case the same as mean_bias) and a regression slope of 2.7, the results indicate that the RK model is overall too warm and exhibits a narrower temperature range, underrepresenting the spatial variability present in the reference data.

```{r Mean Temperature and Correlation by Landuse Class}

# Align raster CRS to land use polygons
mod_h   <- project(mod_h, landuse)
ref30_h <- project(ref30_h, landuse)

# Extract mean temperature per land use polygon
mod_vals <- terra::extract(mod_h, landuse, fun = mean, na.rm = TRUE)
ref_vals <- terra::extract(ref30_h, landuse, fun = mean, na.rm = TRUE)

# Add mean values to land use data
landuse$rk_mean  <- mod_vals[, 2]
landuse$ref_mean <- ref_vals[, 2]

# Calculate model–reference difference
landuse <- landuse %>%
  mutate(diff = rk_mean - ref_mean)

# Summarize by land use class
landuse_summary <- landuse %>%
  st_drop_geometry() %>%
  group_by(class_2018) %>%
  summarise(
    rk_mean   = mean(rk_mean, na.rm = TRUE),
    ref_mean  = mean(ref_mean, na.rm = TRUE),
    diff_mean = mean(diff, na.rm = TRUE),
    corr      = cor(rk_mean, ref_mean, use = "complete.obs")
  )

print(landuse_summary)
```

```{r Color Palette for CORINE Landcover}

# Define color palette for land cover classes based on CORINE
landcover_colors <- c(
  "Continuous Urban fabric (S.L. > 80%)" = "#a40000",
  "Discontinuous Dense Urban Fabric (S.L.: 50% - 80%)" = "#d73027",
  "Discontinuous Medium Density Urban Fabric (S.L.: 30% - 50%)" = "#f46d43",
  "Discontinuous Low Density Urban Fabric (S.L.: 10% - 30%)" = "#fdae61",
  "Discontinuous very Low Density Urban Fabric (S.L.: < 10%)" = "#fee090",
  "Isolated Structures" = "#ffffbf",
  "Industrial, commercial, public, military and private units" = "#bdbdbd",
  "Fast transit roads and associated land" = "#969696",
  "Other roads and associated land" = "#636363",
  "Railways and associated land" = "#252525",
  "Port areas" = "#878787",
  "Airports" = "#cccccc",
  "Mineral extraction and dump sites" = "#ffcc99",
  "Construction sites" = "#ffeda0",
  "Land without current use" = "#f0f0f0",
  "Green urban areas" = "#78c679",
  "Sports and leisure facilities" = "#41ab5d",
  "Arable land (annual crops)" = "#e6ab02",
  "Permanent crops" = "#fee391",
  "Pastures" = "#a1d99b",
  "Complex and mixed cultivation patterns" = "#c2e699",
  "Orchards" = "#addd8e",
  "Forests" = "#006837",
  "Herbaceous vegetation associations" = "#31a354",
  "Open spaces with little or no vegetation" = "#d9f0a3",
  "Wetlands" = "#80cdc1",
  "Water" = "#67a9cf"
)
```

```{r}

# Summarise model–reference correlation by land cover class
cor_summary <- landuse %>%
  st_drop_geometry() %>%
  group_by(class_2018) %>%
  summarise(
    n = n(),
    corr = cor(rk_mean, ref_mean, use = "complete.obs"),
    rk_mean = mean(rk_mean, na.rm = TRUE),
    ref_mean = mean(ref_mean, na.rm = TRUE),
    diff_mean = mean(rk_mean - ref_mean, na.rm = TRUE)
  ) %>%
  arrange(desc(corr))

# remove classes without valid correlations
cor_summary <- cor_summary %>% filter(!is.na(corr))

# Plot correlation by land cover class
p_landcover_corr <- ggplot(cor_summary, aes(
  x = reorder(class_2018, corr), 
  y = corr,                      
  fill = class_2018               
)) +
  geom_col(width = 0.6, color = "grey30") +                         
  geom_text(aes(label = sprintf("%.2f", corr)),         
            hjust = -0.1, size = 3, color = "black") +
  scale_fill_manual(values = landcover_colors) +                
  coord_flip() +                                                    
  geom_hline(yintercept = 0, color = "grey50", linetype = "dashed") + 
  labs(
    title = "Temperature Correlation by Land Cover Class (09 July 2023, 14:00)",
    subtitle = "Pearson’s r between RK model and reference temperatures",
    caption = "Data: RK model output, Urban Climate Analysis Heidelberg (2023),\nUrban Atlas Land Cover/Land Use (2018), Europe, 6-yearly",
    x = NULL,
    y = "Pearson Correlation (r)"
  ) +
  theme_bw(base_size = 13) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14), 
    plot.subtitle = element_text(hjust = 0.5, size = 10, margin = margin(b = 8)), 
    legend.position = "none",                                          
    axis.text.y = element_text(size = 9),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank()
  )

# Display the plots
p_landcover_corr

# Save as PNG
ggsave("../plots/landcover_correlation_2023-07-09_14-00.png", p_landcover_corr, width = 10, height = 5, dpi = 300)
```
