---
title: "Test_kriging"
author: "Johannes Gruenewald"
date: "2024-11-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Databasis for this code is: https://pages.cms.hu-berlin.de/EOL/gcg_quantitative-methods/Lab14_Kriging.html
This is just a test run to get familiar with coding in R and how to run the basic steps which are necessary for a successful interpolation.

```{r}
#install.packages("mapview")

library("gstat")   # geostatistics
library("mapview") # map plot
library("sf")      # spatial vector data
library("stars")   # spatial-temporal data
library("terra")   # raster data handling 
library("ggplot2") # plotting
library(tidyverse)
mapviewOptions(fgb = FALSE)
```


Read temperature points
```{r}
temp <- read_sf('C:/Users/user/OneDrive/Documents/Uni/Master Heidelberg/Masterarbeit/data/temp/subset_temp_data.gpkg')
temp
```

```{r}
mapview(temp['temperature'])
```

Create a grid
```{r}
bbox <- st_bbox(temp)
bbox
```

```{r}
cell_size <- 0.001

x <- seq(bbox$xmin, bbox$xmax, by=cell_size)
y <- seq(bbox$ymin, bbox$ymax, by=cell_size)
```

```{r}
temp_grid <- expand.grid(x=x, y=y)
plot(temp_grid$x, temp_grid$y, pch=19, cex=0.1)
```

```{r}
temp_grid$tmp <- 1
temp_grid <- st_as_stars(temp_grid, crs=st_crs(temp))
st_crs(temp_grid) <- st_crs(temp) # re-assign crs to be safe
```

IDW Interpolation
```{r}
# Remove rows with NA in 'temperature'
temp <- temp[!is.na(temp$temperature), ]

temp.idw <- gstat::idw(temperature ~ 1, locations=temp, newdata=temp_grid, idp = 2)
```

```{r}
temp.idw
```

Plot the predictions
```{r}
plot(rast(temp.idw["var1.pred"]))
plot(temp["temperature"], col=1, cex=0.5, add=T, type="p")
```

Sample Variogram
```{r}
temp.v <- gstat::variogram(log(temperature) ~ 1, temp)
plot(temp.v, ylab=bquote(gamma), xlab=c("h (separation distance in m)"))
```

Direction dependence
```{r}
plot(variogram(log(temperature) ~ 1, temp, alpha = c(0, 45, + 90, 135)))
```
Fit a variogram
```{r}
temp.sph <- vgm(psill=0.03, "Sph", range=3, nugget=0.0002)
plot(temp.v, temp.sph, cutoff=4)
```

Fit variogram
```{r}
temp.vfit <- fit.variogram(temp.v, temp.sph)
temp.vfit
```

```{r}
plot(temp.v, temp.vfit)
```

Ordinary kriging
```{r}
# create sample variogram
#temp.v <- gstat::variogram(log(temperature) ~ 1, temp)

# fit variogram model
#temp.vfit <- gstat::fit.variogram(temp.v, vgm(0.03, "Sph", 3, 0.0002))

# ordinary kriging
lz.ok <- gstat::krige(log(temperature) ~ 1, temp, temp_grid, temp.vfit)
```

Plot results of ordinary kriging
```{r}
plot(rast(lz.ok['var1.pred']))
plot(temp["temperature"], col="blue", cex=0.5, type="p", add=T)
```

Plot variance values of ordinary kriging
```{r}
plot(rast(lz.ok['var1.var']))
plot(temp["temperature"], col="blue", cex=0.5, type="p", add=T)
```

Universal kriging includes other covariates to account for potential trends
```{r}
elev <- terra::rast("../data/hd_elev.tif")
```

```{r}
elev <- terra::project(elev,"EPSG:4326")
crs(elev)
```

```{r}
temp <- temp %>% 
  filter(!is.na(temperature)) %>%
  group_by(entity_id) %>%
  slice_max(order_by = dateobserved, n = 1) %>% 
  dplyr::select(c(temperature))
```

```{r}
# Add X Y Coordinates to data_temp This is necessary for later steps
temp$X <- st_coordinates(temp)[,1]
temp$Y <- st_coordinates(temp)[,2]
```

```{r}
# get elevation at the sites
elevation <- terra::extract(elev,
                             temp,
                             buffer=0)
```

```{r}
# unlist elevations to add as column to data
data_temp$elevation <- elevation[,2]
```

```{r}
plot(data_temp$elevation)
```

